---
title: "Reproducible Research: Peer Assessment 1"
author: William
date: April 13, 2019
output: md_document
---



<header>
Reproducible Research: Peer Assessment 1
============
</header>




```{r setup, include=FALSE}
#a nice preamble
knitr::opts_chunk$set(echo = TRUE)
```
## Initialization
I will begin with all of the packages I intend to use.
I have commented out installation instructions that are implied, but would be useful for first time users.

```{r}
# options(repos = c(CRAN = "https://cran.rstudio.com"))
# install.packages("plyr") 
# install.packages("ggplot2")
# install.packages("HistogramTools")
# install.packages("rmarkdown")
# install.packages("gridExtra")
# library(rmarkdown)
library(plyr)
library(ggplot2)
library(HistogramTools)
library(gridExtra)
```


## Loading and preprocessing the data
Set the working directory and initialize the variables necessary to extract the data for the course.
```{r}
setwd("C:/Users/willi/OneDrive/Documents/work-rosetta/Data Science Academy/05 - Reproducible Research/project 1")
url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
str_zip_filename <- "Exercise Data.zip"
str_input_filename_01<-"activity.csv"

```
Uncomment and run the following section to download the ZIP file and extract its contents.
This typically only needs to happen once.

```{r}
# download.file(url, str_zip_filename, mode = "wb")
# unzip(str_zip_filename, files = NULL)
```

This block loads the input file into our R session for use in the analysis. 
```{r}
dfdata_raw <- read.csv(str_input_filename_01,stringsAsFactors = FALSE)
dfdata <- na.omit(dfdata_raw)
```


## What is mean total number of steps taken per day?
This portion of the code aggregates the total step data at the daily level and provides a quick histogram for visual reasonability check.
```{r}
dailydata<-aggregate(dfdata$steps,by=list(date=dfdata$date),FUN=sum)
hist(dailydata$x,breaks=5,xlab="Steps",main="Total steps per day")
```

The mean number of steps per day is:

```{r}
median(dailydata$x)
```
and the median is:

```{r}
mean(dailydata$x)
```


## What is the average daily activity pattern?
Within a given 5 minute interval, the average number of steps taken across all days contained within the dataset can be seen in the following histogram:

```{r}
intervaldata<-aggregate(dfdata$steps,by=list(interval=dfdata$interval),FUN=mean)
#head(intervaldata)
ggplot(intervaldata,aes(x=interval,y=x)) +
        geom_line() + 
        ylab("Average Number of Steps")+
        xlab("Interval")+
        ggtitle("Average Number of Steps per Interval")
```


The 5 minute interval in which the average number of steps is the highest is
```{r}
intervaldata[intervaldata$x==max(intervaldata$x),1]
```
and during that interval the user took the following number of steps
```{r}
max(intervaldata$x)
```


## Imputing missing values

The following can be used to identify those rows with missing values, and also to identify how many such rows exist:
```{r}
dfdata_na <- dfdata_raw[is.na(dfdata_raw$steps)==TRUE,]
nrow(dfdata_na)
```

To fill these missing values, I will use the average over a given interval which was computed earlier.
To get a better idea of when these null intervals occur, I will attempt to visualize the missing data by day and time on the x and y axes of a plot, respectively.
```{r}
ggplot(dfdata_na,aes(y=interval,x=date)) +
        geom_line()
unique(dfdata_na$date)
```
It appears that there are 8 days which are missing data, so it is reasonable to fill these missing data rows in with an average value.

The following code labels the given data points as either part of the originally filled data, or a null data point that will be filled in.
```{r}
dfdata_na$isna_ind <- TRUE
dfdata$isna_ind <- FALSE
```

This code does the actual filling of the missing values:
```{r}
dfdata_na$steps <- intervaldata[dfdata_na$interval==intervaldata$interval,"interval"]
```

Finally, the filled values can be merged with the original values as follows:
```{r}
na_data_merged<-merge(x = dfdata_na, y=intervaldata, by.x="interval", by.y = "interval")
na_data_merged$steps<-na_data_merged$x
na_data_merged<-within(na_data_merged,rm(x))
na_data_merged<-na_data_merged[,c("steps","date","interval","isna_ind")]

filled_data<-rbind(na_data_merged,dfdata)
filleddailydata<-aggregate(filled_data$steps,by=list(date=filled_data$date),FUN=sum)
```

For a visual representation of the data, reference the following histogram and compare it to the previous version which excluded NAs:
```{r}
#4
hist(filleddailydata$x,breaks=5,xlab="Steps",main="Total steps per day")
```

The new mean and median daily step values have changed slightly after the filling and are:
```{r}
median(filleddailydata$x)
mean(filleddailydata$x)

```

## Are there differences in activity patterns between weekdays and weekends?

To answer the question, first we will reformat the date column in a more useful format and then add a new column containing the name of the day of the week:
```{r}

filled_data$date<-as.Date(as.character(filled_data$date),format ="%Y-%m-%d")
filled_data$dayname<-weekdays(filled_data$date)
```

Next we will add a new column to the dataset indicating whether each row is part of the weekend or a weekday:
```{r}
weekdays<-c("Monday","Friday","Wednesday","Thursday","Tuesday")
filled_data$weekend_ind<-filled_data$dayname %in% weekdays
filled_data$weekend_factor<- factor(filled_data$weekend_ind)
filled_data$weekend_factor<-revalue(filled_data$weekend_factor,c("TRUE" = "Weekday", "FALSE"="Weekend"))
```

With the data factored on weekday vs weekend, we will split it into two data sets for convenience and average each at the daily level:
```{r}
weekdaydata<-filled_data[filled_data$weekend_factor=="Weekday",]
weekenddata<-filled_data[filled_data$weekend_factor=="Weekend",]
weekdayintervaldata<-aggregate(weekdaydata$steps,by=list(interval=weekdaydata$interval),FUN=mean)
weekendintervaldata<-aggregate(weekenddata$steps,by=list(interval=weekenddata$interval),FUN=mean)
```


Finally, we will create two graphs using GGPLOT2 and then display both:
```{r}
p1<-ggplot(weekendintervaldata,aes(x=interval,y=x)) +
        geom_line() + 
        ylab("Average Number of Steps")+
        xlab("Interval")+
        ggtitle("Average Number of Steps per Interval on Weekends")


p2<-ggplot(weekdayintervaldata,aes(x=interval,y=x)) +
        geom_line() + 
        ylab("Average Number of Steps")+
        xlab("Interval")+
        ggtitle("Average Number of Steps per Interval on Weekdays")

grid.arrange(p1,p2,nrow=2)
```

In short, there is a noticeable difference between activity during the week and over the weekend.
